---
title: "R Warm-up for Visualisation"
author: "Priit Adler, Nurlan Kerimov"
date: "2026-04-02"
output: html_document
editor_options:
  markdown:
    wrap: 80
---

```{r setup, include=FALSE}
library(knitr)
library(dplyr)
library(tibble)

opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

cars <- mtcars %>%
  rownames_to_column(var = "car") %>%   # move row names (car models) into a real column
  mutate(
    cyl = factor(cyl),                  # treat cylinder count as a category, not a number
    gear = factor(gear),                # same for number of gears
    am = factor(am, levels = c(0, 1), labels = c("Automatic", "Manual"))  # give 0/1 readable labels
  )
```

## Why this document?

This warm-up is a short prerequisite companion for the ggplot2 session. Use it
to refresh the small data operations that often happen just before plotting.

This instructor version keeps the same structure as the learner handout, but
includes complete solutions, small expected outputs, and short teaching notes.

## Course preconditions

Before coming to the visualisation course, learners should be comfortable with:

- basic R and RStudio navigation
- running code chunks in R Markdown
- the idea of tidy data: rows are observations and columns are variables
- the difference between categorical and continuous variables
- factors as stored categories with labels or a specific order

If these ideas are still new, work through this document first and keep it open
during the plotting session.

## Meet the dataset

In this warm-up we use the built-in `mtcars` dataset for every example.

```{r}
glimpse(cars)
```

### Data Dictionary: mtcars Features

The `mtcars` dataset contains 32 observations of cars (rows) with 11 features (columns) collected from 1974 Motor Trend magazine:

**Continuous Variables (Numeric measurements):**

- **`mpg`**: Miles per gallon - fuel efficiency of the vehicle
- **`disp`**: Displacement in cubic inches - the total volume of all cylinders in the engine
- **`hp`**: Gross horsepower - the raw power output of the engine
- **`drat`**: Rear axle ratio - affects acceleration and fuel efficiency
- **`wt`**: Weight in 1000 lbs - the mass of the vehicle
- **`qsec`**: 1/4 mile time in seconds - acceleration performance (lower is faster)

**Categorical Variables (Categories turned into factors):**

- **`cyl`**: Number of cylinders (4, 6, or 8) - affects power and fuel efficiency
- **`vs`**: Engine type (0 = V-shaped, 1 = straight-line) - engine configuration
- **`am`**: Transmission type (0 = Automatic, 1 = Manual) - how power is transferred
- **`gear`**: Number of forward gears (3, 4, or 5) - transmission gears

**Other:**

- **`carb`**: Number of carburetors - fuel delivery system component
- **`car`**: Car name/model (added during data setup)

A few columns deserve special attention:

- `mpg`, `hp`, `wt`, and `disp` are continuous variables
- `cyl`, `gear`, and `am` are categories, so we turned them into factors

## Exercise 1: `select()`

Pick a small set of columns that would be useful in a fuel-efficiency summary:
`car`, `mpg`, `hp`, `wt`, and `am`.

```{r}
cars %>%
  select(car, mpg, hp, wt, am) %>%  # keep only these five columns, drop the rest
  head()                             # show the first 6 rows (default)
```

Expected result: a 6-row table with only the requested five columns.

Teaching note: this is a good point to remind learners that `select()` changes
columns, not rows.

## Exercise 2: `filter()`

Find manual cars with more than 25 miles per gallon.

```{r}
manual_high_mpg <- cars %>%
  filter(am == "Manual", mpg > 25)  # both conditions must be true: manual AND efficient

manual_high_mpg
```

Expected result: a short table of efficient manual cars such as the Fiat 128,
Honda Civic, and Lotus Europa.

Teaching note: use this moment to connect filtering to later plot subsets, for
example "show only 4-cylinder cars" or "focus on one treatment group."

## Exercise 3: `arrange()`

Sort the data from highest to lowest fuel efficiency. If two cars have similar
fuel efficiency, place the lighter car first.

```{r}
cars %>%
  arrange(desc(mpg), wt) %>%  # desc() reverses order for mpg; wt stays ascending as tiebreaker
  head(10)                     # show only the top 10 rows
```

Expected result: the most fuel-efficient cars appear first, with light cars
breaking ties.

Teaching note: learners often forget that you can mix `desc()` with regular
ascending ordering inside the same `arrange()` call.

## Exercise 4: `mutate()`

Create two new variables:

- `power_to_weight` = `hp / wt`
- `mpg_band` with values `"low"`, `"medium"`, and `"high"`

Use this rule for `mpg_band`:

- `"high"` if `mpg >= 25`
- `"medium"` if `mpg >= 18` and `< 25`
- `"low"` otherwise

```{r}
cars_augmented <- cars %>%
  mutate(
    power_to_weight = hp / wt,        # derived ratio: higher means more powerful for its weight
    mpg_band = case_when(             # case_when works like a multi-condition if/else
      mpg >= 25 ~ "high",             # first matching condition wins
      mpg >= 18 ~ "medium",           # only reached if mpg < 25
      TRUE ~ "low"                    # TRUE is the catch-all fallback (like else)
    )
  )

cars_augmented %>%
  select(car, mpg, hp, wt, power_to_weight, mpg_band) %>%  # narrow down to relevant columns
  head(10)
```

Expected result: the output includes one continuous derived variable and one
categorical variable.

Teaching note: this task is a natural bridge to plotting because learners often
create categories like `mpg_band` for colour or faceting.

## Exercise 5: `group_by()` + `summarise()`

Create a summary table by number of cylinders. Include:

- number of cars
- mean `mpg`
- mean `hp`

```{r}
cars %>%
  group_by(cyl) %>%          # split into groups: one per cylinder count (4, 6, 8)
  summarise(
    n_cars = n(),             # n() counts rows in each group
    mean_mpg = mean(mpg),
    mean_hp = mean(hp)
  )
```

Expected result: one row each for 4, 6, and 8 cylinders.

Teaching note: summarised tables are perfect raw material for bar plots,
line plots, and heatmaps later in the course.

## Exercise 6: one more grouped summary

Create a summary table by transmission and cylinders. This is the kind of table
that later becomes a bar plot, box plot, or heatmap.

Include:

- `am`
- `cyl`
- mean `mpg`
- mean `wt`

```{r}
summary_by_transmission <- cars %>%
  group_by(am, cyl) %>%          # group by two variables: one row per unique am+cyl combination
  summarise(
    mean_mpg = mean(mpg),
    mean_wt = mean(wt)
  )

summary_by_transmission          # print the saved result
```

Expected result: a compact summary table with one row per transmission and
cylinder combination.

Teaching note: point out that this exact shape is what many plotting geoms want
as input.

## Checkpoint

These small verbs show up again and again when building plots:

- `select()` helps you focus on the variables you want to show
- `filter()` limits the data to a relevant subset
- `arrange()` helps you inspect patterns before plotting
- `mutate()` creates variables for labels, colours, groups, or facets
- `group_by()` and `summarise()` create summary tables for bars, lines, and
  heatmaps

In the main visualisation course, we will assume you can read and modify tables
at about this level.
