---
title: "R Warm-up for Visualisation"
author: "Priit Adler, Nurlan Kerimov"
date: "2026-04-02"
output: html_document
editor_options:
  markdown:
    wrap: 80
---

```{r setup, include=FALSE}
library(knitr)
library(dplyr)
library(tibble)

opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

cars <- mtcars %>%
  rownames_to_column(var = "car") %>%
  mutate(
    cyl = factor(cyl),
    gear = factor(gear),
    am = factor(am, levels = c(0, 1), labels = c("Automatic", "Manual"))
  )
```

## Why this document?

This warm-up is a short prerequisite companion for the ggplot2 session. Use it
to refresh the small data operations that often happen just before plotting.

## Course preconditions

Before coming to the visualisation course, learners should be comfortable with:

- basic R and RStudio navigation
- running code chunks in R Markdown
- the idea of tidy data: rows are observations and columns are variables
- the difference between categorical and continuous variables
- factors as stored categories with labels or a specific order

If these ideas are still new, work through this document first and keep it open
during the plotting session.

## Meet the dataset

In this warm-up we use the built-in `mtcars` dataset for every example.

```{r}
glimpse(cars)
```

A few columns deserve special attention:

- `mpg`, `hp`, `wt`, and `disp` are continuous variables
- `cyl`, `gear`, and `am` are categories, so we turned them into factors

## Exercise 1: `select()`

Pick a small set of columns that would be useful in a fuel-efficiency summary:
`car`, `mpg`, `hp`, `wt`, and `am`.

```{r}
# Select the requested columns from cars.
# Print the first 6 rows of the result.
```

## Exercise 2: `filter()`

Find manual cars with more than 25 miles per gallon.

```{r}
# Filter cars so that only manual transmission cars with mpg > 25 remain.
```

## Exercise 3: `arrange()`

Sort the data from highest to lowest fuel efficiency. If two cars have similar
fuel efficiency, place the lighter car first.

```{r}
# Arrange by descending mpg and then ascending wt.
# Show the first 10 rows.
```

## Exercise 4: `mutate()`

Create two new variables:

- `power_to_weight` = `hp / wt`
- `mpg_band` with values `"low"`, `"medium"`, and `"high"`

Use this rule for `mpg_band`:

- `"high"` if `mpg >= 25`
- `"medium"` if `mpg >= 18` and `< 25`
- `"low"` otherwise

```{r}
# Add power_to_weight and mpg_band.
# Keep car, mpg, hp, wt, power_to_weight, and mpg_band in the printed result.
```

## Exercise 5: `group_by()` + `summarise()`

Create a summary table by number of cylinders. Include:

- number of cars
- mean `mpg`
- mean `hp`

```{r}
# Group by cyl and calculate n_cars, mean_mpg, and mean_hp.
```

## Exercise 6: one more grouped summary

Create a summary table by transmission and cylinders. This is the kind of table
that later becomes a bar plot, box plot, or heatmap.

Include:

- `am`
- `cyl`
- mean `mpg`
- mean `wt`

```{r}
# Group by am and cyl, then summarise mean_mpg and mean_wt.
```

## Checkpoint

These small verbs show up again and again when building plots:

- `select()` helps you focus on the variables you want to show
- `filter()` limits the data to a relevant subset
- `arrange()` helps you inspect patterns before plotting
- `mutate()` creates variables for labels, colours, groups, or facets
- `group_by()` and `summarise()` create summary tables for bars, lines, and
  heatmaps

In the main visualisation course, we will assume you can read and modify tables
at about this level.
