---
title: "R Visualisation with ggplot2"
author: "Priit Adler, Nurlan Kerimov"
date: "2026-04-02"
output: html_document
editor_options:
  markdown:
    wrap: 80
---

```{r setup, include=FALSE}
library(knitr)
library(ggplot2)
library(dplyr)
library(patchwork)
library(tibble)

opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

cars <- mtcars %>%
  rownames_to_column(var = "car") %>%
  mutate(
    cyl = factor(cyl),
    gear = factor(gear),
    am = factor(am, levels = c(0, 1), labels = c("Automatic", "Manual"))
  )
```

## Course scope

This course focuses on visualisation with `ggplot2`. We use one dataset from
start to finish so that the plotting ideas stay in focus instead of constantly
switching contexts.

## Prerequisites

Before starting this session, learners should already be comfortable with:

- basic R and RStudio use
- running code chunks in R Markdown
- tidy data
- factors
- categorical and continuous variables
- small table operations such as `select()`, `filter()`, `mutate()`,
  `group_by()`, and `summarise()`

If you want a quick refresher on those ideas, use the `2026-04-02-R-warmup.Rmd`
document first.

## Meet the dataset

We use the built-in `mtcars` dataset throughout the course.

```{r}
glimpse(cars)
```

The main variables we will reuse are:

- `mpg`: miles per gallon
- `wt`: car weight in 1000 lb
- `hp`: horsepower
- `cyl`, `gear`, and `am`: categories that are useful for grouping and colour

## Your first ggplot

`ggplot2` builds plots step by step. We start by giving it data, mapping
variables to aesthetics, and adding a geom.

```{r}
ggplot(cars, aes(x = wt, y = mpg)) +
  geom_point(size = 3, color = "steelblue")
```

## Grammar of graphics

The grammar of graphics breaks a plot into a few reusable pieces:

- data: the table used in the plot
- aesthetics: how variables are mapped to visual properties such as x, y, and
  colour
- geoms: the visible marks such as points, bars, or boxes
- scales: how data values are translated into positions, colours, or sizes
- themes: the non-data styling around the plot

Because each part is separate, we can keep adding or changing layers without
starting again from scratch.

## Aesthetics

Aesthetics connect variables in the dataset to what we see on screen. In
practice, this means asking questions like:

- which variable goes on the x-axis?
- which variable goes on the y-axis?
- should colour represent a category or a number?

### Mapping versus setting

When an aesthetic belongs to the data, put it inside `aes()`. When you want a
fixed value, set it outside `aes()`.

```{r fig.width=10, fig.height=4}
mapped_plot <- ggplot(cars, aes(x = wt, y = mpg, color = am)) +
  geom_point(size = 3) +
  labs(title = "Colour mapped to a variable", color = "Transmission")

set_plot <- ggplot(cars, aes(x = wt, y = mpg)) +
  geom_point(size = 3, color = "steelblue") +
  labs(title = "Colour set to a fixed value")

mapped_plot | set_plot
```

You can also map more than one aesthetic at the same time.

```{r}
ggplot(cars, aes(x = wt, y = mpg, color = am, shape = cyl)) +
  geom_point(size = 3)
```

### Task

Create a scatter plot of `disp` vs `hp`, map `color` to `cyl`, and set point
size to `3`.

```{r}
# Build the plot here.
```

## Geoms

Different geoms are useful for different variable types. With the same dataset,
we can make several valid plots by changing only the geom.

```{r fig.width=10, fig.height=8}
scatter_geom <- ggplot(cars, aes(x = wt, y = mpg)) +
  geom_point(color = "steelblue", size = 3) +
  labs(title = "geom_point()")

hist_geom <- ggplot(cars, aes(x = mpg)) +
  geom_histogram(binwidth = 2, fill = "tomato", color = "white") +
  labs(title = "geom_histogram()")

bar_geom <- ggplot(cars, aes(x = cyl)) +
  geom_bar(fill = "grey40") +
  labs(title = "geom_bar()", x = "Cylinders")

box_geom <- ggplot(cars, aes(x = cyl, y = mpg)) +
  geom_boxplot(fill = "khaki") +
  labs(title = "geom_boxplot()", x = "Cylinders")

(scatter_geom | hist_geom) / (bar_geom | box_geom)
```

As a rule of thumb:

- use scatter plots for two continuous variables
- use histograms for one continuous variable
- use bar plots for counts of categories
- use box plots when comparing a continuous variable across categories

## Layers and position adjustments

One of the strengths of `ggplot2` is that you can add more than one layer to
the same plot.

### Box plot with jittered points

```{r}
ggplot(cars, aes(x = cyl, y = mpg, fill = cyl)) +
  geom_boxplot(alpha = 0.5, outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.75, size = 2) +
  labs(x = "Cylinders", y = "Miles per gallon", fill = "Cylinders")
```

The box plot shows the summary distribution, while the jittered points reveal
the individual cars.

### Scatter plot with a trend line and labels

```{r}
ggplot(cars, aes(x = wt, y = mpg)) +
  geom_point(aes(color = am), size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "black", linewidth = 0.8) +
  geom_text(
    data = cars %>% filter(mpg > 30 | wt > 5),
    aes(label = car),
    check_overlap = TRUE,
    nudge_y = 1,
    size = 3
  ) +
  labs(color = "Transmission")
```

### Task

Build a plot of `mpg` by `gear` that combines `geom_boxplot()` and
`geom_jitter()`. Colour the points by `am`.

```{r}
# Build the layered plot here.
```

## Scales

Scales control how values are translated into positions, colours, fills, and
labels.

### Numeric scales

Numeric scales are useful when you want cleaner breaks or more readable axes.

```{r fig.width=10, fig.height=4}
numeric_default <- ggplot(cars, aes(x = hp, y = mpg)) +
  geom_point(size = 3, color = "grey40") +
  labs(title = "Default numeric scales")

numeric_scaled <- ggplot(cars, aes(x = hp, y = mpg)) +
  geom_point(size = 3, color = "grey40") +
  scale_x_continuous(breaks = seq(50, 350, by = 50)) +
  scale_y_continuous(breaks = seq(10, 35, by = 5)) +
  labs(title = "Custom numeric breaks")

numeric_default | numeric_scaled
```

### Discrete scales and factor order

Ordering categories can change the story a plot tells. Here the first box plot
uses the default factor order, while the second reorders cylinders by median
`mpg`.

```{r fig.width=10, fig.height=4}
default_order_plot <- ggplot(cars, aes(x = factor(cyl), y = mpg)) +
  geom_boxplot(fill = "grey80") +
  labs(title = "Default order", x = "Cylinders")

reordered_plot <- ggplot(cars, aes(x = reorder(factor(cyl), mpg, median), y = mpg)) +
  geom_boxplot(fill = "tomato") +
  labs(title = "Reordered by median mpg", x = "Cylinders")

default_order_plot | reordered_plot
```

This is the point where factor handling matters directly for visualisation.

### Colour scales and fill scales

Use `color` for points and lines. Use `fill` for objects with an inside area,
such as bars, tiles, and box plots.

```{r fig.width=10, fig.height=4}
color_scale_plot <- ggplot(cars, aes(x = wt, y = mpg, color = hp)) +
  geom_point(size = 4) +
  scale_color_gradient(low = "skyblue", high = "firebrick") +
  labs(title = "Continuous colour scale", color = "Horsepower")

mean_mpg_by_cyl_gear <- cars %>%
  group_by(cyl, gear) %>%
  summarise(mean_mpg = mean(mpg), .groups = "drop")

fill_scale_plot <- ggplot(mean_mpg_by_cyl_gear, aes(x = cyl, y = gear, fill = mean_mpg)) +
  geom_tile(color = "white", linewidth = 0.8) +
  geom_text(aes(label = round(mean_mpg, 1)), color = "white", fontface = "bold") +
  scale_fill_gradient(low = "khaki1", high = "darkgreen") +
  labs(
    title = "Continuous fill scale",
    x = "Cylinders",
    y = "Gears",
    fill = "Mean mpg"
  )

color_scale_plot | fill_scale_plot
```

## Titles, labels, limits, and themes

Polishing matters. A plot becomes easier to read when it has a clear title,
useful axis labels, and a consistent theme.

```{r}
styled_plot <- ggplot(cars, aes(x = wt, y = mpg, color = am)) +
  geom_point(size = 3) +
  labs(
    title = "Heavier cars tend to have lower fuel efficiency",
    subtitle = "Transmission highlights one part of the pattern in mtcars",
    x = "Weight (1000 lb)",
    y = "Miles per gallon",
    color = "Transmission"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")

styled_plot
```

If you want to zoom into a region without dropping observations, use
`coord_cartesian()`.

```{r}
styled_plot +
  coord_cartesian(xlim = c(1.5, 4.5), ylim = c(10, 35)) +
  labs(caption = "Zoomed with coord_cartesian()")
```

## Exporting plots with `ggsave()`

`ggsave()` writes a plot to a file. In this example we save to temporary files
so that knitting the document does not clutter the course folder.

```{r}
png_file <- tempfile("mtcars-plot-", fileext = ".png")
pdf_file <- tempfile("mtcars-plot-", fileext = ".pdf")

ggsave(filename = png_file, plot = styled_plot, width = 8, height = 5, dpi = 300)
ggsave(filename = pdf_file, plot = styled_plot, width = 8, height = 5)

png_file
pdf_file
```

When you make your own report, replace `tempfile()` with a real filename such as
`"my-figure.png"`.

## Combining plots with patchwork

The `patchwork` package makes it easy to combine several ggplots into one
layout.

```{r fig.width=11, fig.height=8}
dashboard_plot <- (styled_plot / (reordered_plot | fill_scale_plot)) +
  plot_annotation(
    title = "Three views of the mtcars dataset",
    subtitle = "Scatter plot, reordered box plot, and summary heatmap",
    caption = "Data source: mtcars"
  ) +
  plot_layout(guides = "collect")

dashboard_plot
```

## Final task

Create a two-panel figure about fuel efficiency in `mtcars`.

Requirements:

- one panel should use two continuous variables
- one panel should use a categorical x-axis
- choose at least one scale intentionally
- add meaningful labels or a title

```{r fig.width=10, fig.height=6}
# Build your final figure here.
```
